---
title: "Assignment 7 - Answers"
author: "Charles Lang"
date: "11/30/2016"
output: html_document
---

In the following assignment you will be looking at data from an one level of an online geography tutoring system used by 5th grade students. The game involves a pre-test of geography knowledge (pre.test), a series of assignments for which you have the average score (av.assignment.score),  the number of messages sent by each student to other students about the assignments (messages), the number of forum posts students posted asking questions about the assignment (forum.posts), a post test at the end of the level (post.test) and whether or not the system allowed the students to go on to the next level (level.up).  

#Upload data
```{r}
D1 <- read.csv("D:/education & psychology/learning analytics/HUDK 4050 Core Mthds Educ Data Mining_Lang/assignments/assignment7-master/online.data.csv", header = TRUE)
```

#Visualization 
```{r}
#histograms of the distributions for all variables
library(plyr)
library(psych)
multi.hist(D1[,sapply(D1, is.numeric)])

#visualize the relationships between variables
pairs(D1[, 2:7])
#capture an intution about the data and the relationships
```

It seems that most pairwise variables have a clear linear relationship. Those three variables that can separate level.up rather well are post.test.score, forum.posts, and messages.

#Classification tree
```{r}
#Create a classification tree using three variables 
library(rpart)
rp <- rpart(factor(level.up)~post.test.score + messages + forum.posts, method="class", data=D1)
printcp(rp)
#Plot and generate a CP table for your tree 
post(rp, file='',title='Tree1')
#probability that a student levels up based on classification tree 
D1$pred <- predict(rp, type = "prob")[,2]

#ROC curve
library(ROCR)
#Plot curve
pred.detail <- prediction(D1$pred, D1$level.up) 
plot(performance(pred.detail, "tpr", "fpr"))
abline(0, 1, lty = 2)
#Calculate AUC
unlist(slot(performance(pred.detail,"auc"), "y.values"))

#use variables not used, compare plots & results. Which model was better? Why?
rp2 <- rpart(level.up~pre.test.score + av.assignment.score, method="class", data=D1)
printcp(rp2) 
post(rp2, file='',title='Tree2') # plot
D1$pred2 <- predict(rp2, type = "prob")[,2]
pred.detail2 <- prediction(D1$pred2, D1$level.up) 
plot(performance(pred.detail2, "tpr", "fpr"))
unlist(slot(performance(pred.detail2,"auc"), "y.values")) # results
```

Based on AUC, the second model is better than the first one, with 0.977 vs. 0.855. The result is also reflected in the ROC plot, with the area of the first model larger than the second one. However, the first model achieved perfection, a rather weird result.

#Thresholds
```{r}
#Find balance threshold for 1st model, generate a variable that classifies accordingly
D1$level.up <- ifelse(D1$level.up=="yes", 1, 0)

diagnostics <- function(thr){
    pred <- sum(D1$pred > thr)
    #diagnostics
    accuracy <- pred / nrow(D1)
    false.pos <- sum(D1$pred>thr & D1$level.up==0)
    precision <- pred / (pred + false.pos)
    false.neg <- sum(D1$pred<=thr & D1$level.up==1)
    recall <- pred/(pred + false.neg)
    #Kappa
    D1$pred1 <- D1$pred > thr
    table1 <- table(D1$level.up, D1$pred1)
    matrix1 <- as.matrix(table1)
    kappa = kappa(matrix1, exact = TRUE)/kappa(matrix1)
    return(c(paste("accuracy: ", accuracy), paste("precision: ", precision), 
             paste("recall: ",  recall), paste("kappa: ", kappa)))
}

diagnostics(.99)
diagnostics(.9)
diagnostics(.8)
diagnostics(.7)
diagnostics(.6)
diagnostics(.05)
diagnostics(.01)
```

Conclusions?
A threshhold of .6 and .7 seems to be the best considering all diagnostic measures. For values over .8, kappa is 1, but the other diagnostics are either 0 or not attainable. For values around .6 and .7, kappa is close to .941, accuracy is around .5 and precision is around .8. For values below 1, kappa is 1, both acccuracy and recall are 1, and precision is .625. Second, kappa does not seem to be a good measure, since in this case all values of the threshold give high kappa values.



